# Spatial data in R {#geodata}
<!-- # Geographic data in R {#geodata} -->
<!-- to discuss -->

\index{geodata}

## Introduction

Two basic models to represent spatial data are vector and raster data model.
<!-- - few introduction sections -->
<!-- - mention GDAL, PROJ, and GEOS -->
<!-- - references to the next sections -->
<!-- - maybe also references to some books (either here or in the next section or both) -->
<!-- - maybe also mention some data sources -->

## Vector data

<!-- - one/two intro sentences -->
<!-- - including r packages used for vector representation -->

### Vector data model

...
The Vector data model is consists of two main elements: geometries and attributes.
The role of geometry is to describe the location and shape of spatial objects.
There are three basic types of geometries: points, lines, and polygons.
All of them are build using the same main idea of coordinates.
A point is represented by a pair of coordinates, usually described as X and Y.
It allows for locating this point in some space.
<!-- short CRS intro -->
X and Y could be unitless, in degrees, or in some measure units, such as meters.
<!-- maybe ref to CRS section here -->
<!-- ways to adjust points aesthetics: point sizes, colors, shapes, (markers/images) -->
<!-- examples of points -->
Points can represents features on different scales, from a GPS position, location of a bench in a park, to a city in a small scale map.
<!-- Points can be also used to represent abstract features -->
A line extends the idea of a point.
It consists of several points (with coordinates)<!--vertex--> that are arranged in some order.
Consecutive points are connected by straight lines.
Therefore, a straight spatial line consists of two points (two pairs of coordinates), while complex spatial lines could be created based on a large number of points.<!--to rewrite-->
It gives the illusion that the line is curved. 
<!-- ways to adjust lines aesthetics: colors, lwd (line width) -->
<!-- in theory lty could be also used - but it is not implemented in tmap -->
<!-- examples of lines -->
Lines are used to represent linear features, such as roads, rivers, boundaries, footpaths, etc. 
Polygon is again a set of ordered points connected by straight lines. 
The only difference is the first and the last point in a polygon has the same coordinates, and thus close the object.
<!-- ways to adjust polygons aesthetics: (fill) colors, (line/borders) colors -->
<!-- examples of polygons -->
Polygons also have one unique feature - they could have holes. 
A polygon hole represents an area inside of the polygon but does not belong to it.
For example, a lake with an island can be depicted as a polygon with a hole.

```{r, echo=FALSE}
library(sf)
library(tmap)
point_data = st_sf(
    id = as.factor(c(1, 2)),
    geom = st_sfc(
      st_multipoint(rbind(c(1, 2))),
      st_multipoint(rbind(c(1.4, 1.2), c(2, 1.5))),
      crs = 4326)
  )
tm_shape(point_data) +
  tm_grid() + 
  tm_symbols(col = "id") +
  tm_layout(main.title = "POINT",
            legend.show = FALSE,
            inner.margins = 0.1)
```

```{r, echo=FALSE}
linestring_sfg1 = st_linestring(rbind(c(0.9, 1.9), c(1.6, 1.9), c(1.8, 1.8)))
linestring_sfg2 = st_linestring(rbind(c(0.8, 1.6), c(1.4, 1.5), c(2, 1.6)))
linestring_sfg3 = st_linestring(rbind(c(1.1, 1), c(1.6, 1.2), c(1.9, 1.2)))
lines_data = st_sf(
  id = as.factor(1:3),
  geom = st_sfc(linestring_sfg1, 
                linestring_sfg2,
                linestring_sfg3,
                crs = 4326)
  )
lines_data_p = st_cast(lines_data, "MULTIPOINT")

tm_shape(lines_data) +
  tm_grid() + 
  tm_lines(col = "id", lwd = 7) +
  tm_shape(lines_data_p) +
  tm_dots(size = 1) +
  tm_layout(main.title = "LINESTRING",
            legend.show = FALSE,
            inner.margins = 0.1)
```

```{r, echo=FALSE}
polygon_sfg1 = st_polygon(list(rbind(c(1.4, 1.6),
                                     c(1.6, 1.6), 
                                     c(1.6, 1.8),
                                     c(1.4, 1.8), 
                                     c(1.4, 1.6))))

polygon_sfg2 = st_multipolygon(list(list(rbind(
  c(1.6, 1.3), c(1.8, 1.45), c(1.6, 1.45), c(1.6, 1.3)
)),
st_polygon(
  list(rbind(
    c(0.9, 1.1), c(1.2, 1.1), c(1.2, 1.4), c(0.9, 1.4), c(0.9, 1.1)
  ),
  rbind(
    c(1.0, 1.2), c(1.1, 1.3), c(1.1, 1.2), c(1.0, 1.2)
  ))
)))

polygon_data = st_sf(
  id = as.factor(1:2),
  geom = st_sfc(polygon_sfg1, 
                polygon_sfg2,
                crs = 4326)
  )

polygon_data_p = st_cast(polygon_data, "MULTIPOINT")

tm_shape(polygon_data) +
  tm_grid() + 
  tm_polygons(col = "id", lwd = 7) +
  tm_shape(polygon_data_p) +
  tm_dots(size = 1) +
  tm_layout(main.title = "POLYGON",
            legend.show = FALSE,
            inner.margins = 0.15)
```

```{r, echo=FALSE}
# tmap_mode("plot")
all_data = rbind(point_data, lines_data, polygon_data)
tm_shape(polygon_data, bbox = all_data) +
    tm_grid() +
    tm_polygons(col = "id", lwd = 7) +
    tm_shape(lines_data) +
    tm_lines(col = "id") +
    tm_shape(point_data) +
    tm_symbols(col = "id") +
  tm_layout(inner.margins = 0.15)
```


<!-- additional dimensions: -->
<!-- - more than two coordinates (XYZM) -->
<!-- - multiobjects -->
<!-- - additional geometries -->

<!-- - what is the vector data model (point coordinates) -->
<!-- - examples -->
<!-- - what's the simple features standard -->
<!-- - main geometry types -->
<!-- - relation between geometries and attributes -->
<!-- - vector file formats -->
<!-- - advantages/disadvantages -->
<!-- - example figure (similar to the one in geocompr, but made with tmap) -->

### The sf package

<!-- "Vector data provide a way to represent real world features" -->
<!-- - how the sf objects are organized -->
<!-- - how to read sf objects from files -->
<!-- - where to find info on how to operate on sf objects -->
<!-- - https://geocompr.github.io/ -->
<!-- - vector simplification? -->
<!-- - stars proxy -->

## Raster data

<!-- one/two intro sentences -->
<!-- including r packages used for raster representation -->

### Raster data model

<!-- - raster data model (grid) -->
<!-- - contionous and categorical rasters -->
<!-- - examples -->
<!-- - single layer rasters vs multilayer rasters -->
<!-- - storing bands vs attributes (either here or in the next section) (data cubes) -->
<!-- - rbg rasters -->
<!-- - regular, rotated, sheared, rectilinear and curvilinear rasters -->
<!-- - raster file formats -->

### The stars package

<!-- - how the stars objects are organized -->
<!-- - how to read stars objects from files -->
<!-- - including reading chunks, changing resolution, and selecting bands -->
<!-- - where to find info on how to operate on stars objects -->
<!-- - https://r-spatial.github.io/stars/index.html -->
<!-- - advice: sometimes/often it is better to prepare spatial object before the mapping, than trying to over-customize the map -->
<!-- - stars proxy -->


## CRS
<!-- mtennekes part -->