# **tmap** in a nutshell {#nutshell}

The **tmap** package allows the creation of thematic maps with great flexibility.
It accepts spatial data in various formats shape objects (section \@ref(shape-objects))
Next, the data can be used to create simple, quick maps (section \@ref(quick-maps)) and more complex and expandable maps (section \@ref(regular-maps)).
These maps can be presented in two modes - as a static map and an interactive one.
Additionally, **tmap** makes it possible to create small multiples map (section \@ref(sm-section)) and map animations (section \@ref(ani-section)).

## Shape objects

<!-- two main data models -->
<!-- many spatial data formats -->
<!-- also many spatial data classes in r -->
<!-- tmap collectively calls all of them shape objects -->
<!-- ref to data chapter  (see chapter \@ref(geodata)) -->

<!-- for example, below, we read ... -->

```{r, message=FALSE}
library(tmap)
library(sf)
library(stars)
ei_elev = read_stars("data/easter_island/ei_elev.tif")
ei_borders = read_sf("data/easter_island/ei_border.gpkg")
ei_roads = read_sf("data/easter_island/ei_roads.gpkg")
ei_points = read_sf("data/easter_island/ei_points.gpkg")
volcanos = subset(ei_points, type == "volcano")
```

<!-- ref also to data appendix -->

<!-- spatial data, no matter the format, usually stores spatial geometries/locations/shapes and related data/attributes -->
<!-- if we just want to present data attributes, we could use regular plotting functions (e.g. ...) or packages (e.g. ) -->
<!-- however, when we create maps - our goal is usually to either just visualize spatial shapes or spatial shapes together with their attributes -->

## Quick maps

<!-- customization vs quick map -->
<!-- exploration vs communication -->
```{r, eval=FALSE, echo=FALSE}
qtm(ei_elev)
```

```{r}
qtm(volcanos)
```

```{r}
qtm(volcanos, symbols.shape = 24, symbols.size = "elevation", title = "Volcanos")
```

## Regular maps

<!-- mention tmap elements -->

<!-- reproduce the above plot -->
```{r, eval=FALSE}
tm_shape(volcanos) +
  tm_symbols(shape = 24, size = "elevation") +
  tm_layout(title = "Volcanos")
```


<!-- add a complete map code -->
<!-- - Layered approach (grammar of graphics) -->
<!-- explain line by line -->
<!-- ref to other parts of the book -->

```{r, warning=FALSE}
my_map = tm_shape(ei_elev) +
  tm_graticules() +
  tm_raster(style = "cont", title = "Elevation (m asl)",
            palette = "-RdYlGn", midpoint = NA) +
  tm_shape(ei_borders) + 
  tm_borders() +
  tm_shape(ei_roads) + 
  tm_lines(lwd = "strokelwd", legend.lwd.show = FALSE) +
  tm_shape(volcanos) +
  tm_symbols(shape = 24, size = "elevation",
             title.size = "Volcanos (m asl)") +
  tm_add_legend(type = "line", title = "Roads", col = "black") +
  tm_compass(position = c("right", "top")) +
  tm_scale_bar() +
  tm_layout(main.title = "Easter Island",
            bg.color = "lightblue")
```

```{r, fig.height=9, fig.asp=0.73}
my_map
```

<!-- refs  -->


## Map modes

<!-- static maps, default -->
<!-- interactive maps -->
```{r}
tmap_mode("view")
```

```{r, eval=FALSE}
my_map
```

```{r}
tmap_mode("plot")
```

```{r, eval=FALSE}
my_map
```

## Small multiples {#sm-section}

## Animations {#ani-section}
