# Save maps {#save}

Maps created programmatically can serve several purposes, from exploratory, through visualizations of the processing steps, to being the final outputs of a given project.
Therefore, often we want just to see our map on the screen, but sometimes we also want to save our map results to an external file.
**tmap** objects can be directly saved to output files with `tmap_save()`^[Standard R approach of saving graphic files by opening a graphic device, e.g., `png()`, plotting the data, and then closing the device with `dev.off()` also works.].

The `tmap_save()` function allows to save our map in three groups of file formats, (a) vector graphics (section \@ref(vector-graphic-formats)), (b) raster graphics (section \@ref(raster-graphic-formats)), and (c) interactive ones (section \@ref(interactive-format)).
Additionally, chapter \@ref(animations) shows how to save map animations with the use of the `tmap_animation()` function.

For the examples in this section, we will use a simple map of the Easter Island polygon with the island name superimposed (not shown).

```{r, message=FALSE, fig.show='hide'}
library(tmap)
library(sf)
ei_borders = read_sf("data/easter_island/ei_border.gpkg")
tm = tm_shape(ei_borders) +
  tm_polygons() +
  tm_text("name", fontface = "italic")
tm
```

## Vector graphic formats

<!-- saving workflow - create tmap object -> tmap save -->

```{r}
tmap_save(tm, "my_map0.png")
```

```{r}
tmap_save(tm, "my_map.png", width = 1000, height = 750, dpi = 300)
```
<!-- units -->
<!-- , asp = 0 -->
<!-- scale -->

```{r}
tmap_save(tm, "my_map.png", device = ragg::agg_png)
```

## Raster graphic formats

```{r}
tmap_save(tm, "my_map.svg")
```

<!-- scale -->
<!-- ref to the fonts section -->

## Interactive format

```{r}
tmap_save(tm, "my_map.html")
```

<!-- add.titles -->
<!-- in.iframe -->
<!-- selfcontained -->
<!-- arguments passed on to device functions or to saveWidget or saveWidgetframe -->